---
title: "clark_blackshaw_reanalysis"
output: html_document
date: "2024-06-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This reference was created from developmental dataset from Clark et al. 2019 (https://doi.org/10.1016/j.neuron.2019.04.010). 
Filtered feature matrices were downloaded for each individual timepoint, merged, and object processed in same manner as datasets generated for this paper.
Celltypes were manually reannotated based on analysis.
Final r object produced here was used a reference object for Seurat reference mapping for celltypes in E16 and P0 miR-9-2 KO datasets

```{r load libraries}
library(Seurat)
library(SeuratData)
library(sctransform)
library(patchwork)
library(dplyr)
library(ggplot2)
library(harmony)
```

```{r Adjust the maximum size of global objects}
# (this may need to be increased later)
options(future.globals.maxSize = 200 * 1024^3)
```

```{r Read the 10x files}
E11.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E11_Clark/outs/filtered_feature_bc_matrix/")
E12.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E12_Clark/outs/filtered_feature_bc_matrix/")
E14.1.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E14_rep1_Clark/outs/filtered_feature_bc_matrix/")
E14.2.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E14_rep2_Clark/outs/filtered_feature_bc_matrix/")
E16.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E16_Clark/outs/filtered_feature_bc_matrix/")
E18.2.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E18_rep2_Clark/outs/filtered_feature_bc_matrix/")
E18.3.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_E18_rep3_Clark/outs/filtered_feature_bc_matrix/")
P0.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_P0_Clark/outs/filtered_feature_bc_matrix/")
P2.2.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_P2_rep2_Clark/outs/filtered_feature_bc_matrix/") # only using one rep of P2 data because it looks like identical datasets were uploaded
P5.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_P5_Clark/outs/filtered_feature_bc_matrix/")
P8.1.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_P8_rep1_Clark/outs/filtered_feature_bc_matrix/") # only using one of two P8 replicates -- they look identical
P14.data <- Read10X(data.dir = "/Clark_Blackshaw_reanalysis/Alignments/count_P14_Clark/outs/filtered_feature_bc_matrix/")
```

```{r Create Seurat objects}
E11 <- CreateSeuratObject(counts = E11.data, project = "E11", min.cells = 3, min.features = 200)
E12 <- CreateSeuratObject(counts = E12.data, project = "E12", min.cells = 3, min.features = 200)
E14.1 <- CreateSeuratObject(counts = E14.1.data, project = "E14.1", min.cells = 3, min.features = 200)
E14.2 <- CreateSeuratObject(counts = E14.2.data, project = "E14.2", min.cells = 3, min.features = 200)
E16 <- CreateSeuratObject(counts =  E16.data, project = "E16", min.cells = 3, min.features = 200)
E18.2 <- CreateSeuratObject(counts = E18.2.data, project = "E18.2", min.cells = 3, min.features = 200)
E18.3 <- CreateSeuratObject(counts =  E18.3.data, project = "E18.3", min.cells = 3, min.features = 200)
P0 <- CreateSeuratObject(counts = P0.data, project = "P0", min.cells = 3, min.features = 200)
P2.2 <- CreateSeuratObject(counts = P2.2.data, project = "P2.2", min.cells = 3, min.features = 200)
P5 <- CreateSeuratObject(counts = P5.data, project = "P5", min.cells = 3, min.features = 200)
P8.1 <- CreateSeuratObject(counts = P8.1.data, project = "P8.1", min.cells = 3, min.features = 200) # only use one of the P8 replicates -- they look identical
P14 <- CreateSeuratObject(counts = P14.data, project = "P14", min.cells = 3, min.features = 200)
```

```{r Add mitochondrial counts}
E11[["percent.mt"]] <- PercentageFeatureSet(E11, pattern = "^mt-")
E12[["percent.mt"]] <- PercentageFeatureSet(E12, pattern = "^mt-")
E14.1[["percent.mt"]] <- PercentageFeatureSet(E14.1, pattern = "^mt-")
E14.2[["percent.mt"]] <- PercentageFeatureSet(E14.2, pattern = "^mt-")
E16[["percent.mt"]] <- PercentageFeatureSet(E16, pattern = "^mt-")
E18.2[["percent.mt"]] <- PercentageFeatureSet(E18.2, pattern = "^mt-")
E18.3[["percent.mt"]] <- PercentageFeatureSet(E18.3, pattern = "^mt-")
P0[["percent.mt"]] <- PercentageFeatureSet(P0, pattern = "^mt-")
P2.2[["percent.mt"]] <- PercentageFeatureSet(P2.2, pattern = "^mt-")
P5[["percent.mt"]] <- PercentageFeatureSet(P5, pattern = "^mt-")
P8.1[["percent.mt"]] <- PercentageFeatureSet(P8.1, pattern = "^mt-")
P14[["percent.mt"]] <- PercentageFeatureSet(P14, pattern = "^mt-")
```

```{r Create violin plots for QC}
VlnPlot(E11, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E11_qc.pdf")

VlnPlot(E12, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E12_qc.pdf")

VlnPlot(E14.1, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E14.1_qc.pdf")

VlnPlot(E14.2, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E14.2_qc.pdf")

VlnPlot(E16, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E16_qc.pdf")

VlnPlot(E18.2, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E18.2_qc.pdf")

VlnPlot(E18.3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "E18.3_qc.pdf")

VlnPlot(P0, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "P0_qc.pdf")

VlnPlot(P2.2, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "P2_qc.pdf")

VlnPlot(P5, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "P5_qc.pdf")

VlnPlot(P8.1, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "P8.1_qc.pdf")

VlnPlot(P8.2, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "P8.2_qc.pdf")

VlnPlot(P14, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= "P14_qc.pdf")
```


```{r Subset the data based on QC metric cutoffs}
# clark, blackshaw data is single cell not single nucleus so using a 10% mitochondrial dna cutoff
E11 <- subset(E11, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
E12 <- subset(E12, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
E14.1 <- subset(E14.1, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
E14.2 <- subset(E14.2, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
E16 <- subset(E16, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
E18.2 <- subset(E18.2, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
E18.3 <- subset(E18.3, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
P0 <- subset(P0, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
P2.2 <- subset(P2.2, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
P5 <- subset(P5, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
P8.1 <- subset(P8.1, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)
P14 <- subset(P14, subset = nCount_RNA > 300 & nCount_RNA < 30000 & percent.mt < 10)

samples <- c("E11", "E12", "E14.1", "E14.2", "E16", "E18.2", "E18.3", "P0", "P2.2", "P5", "P8.1", "P14")

for (sample in samples){
  obj <- get(sample)
  saveRDS(obj, file = paste0(sample, "_unprocessed.rds"))
}
```

```{r merge timepoint replicates}
E14 = merge(E14.1, y = E14.2, add.cell.ids = c("E14.1", "E14.2"))
E18 = merge(E18.2, y = E18.3 , add.cell.ids = c("E18.2", "E18.3"))

samples <- c("E14", "E18")

for (sample in samples){
  obj <- get(sample)

  saveRDS(obj, file = paste0(sample, "_unprocessed.rds"))
}
```

# Create reference for all retinal timepoints

```{r reload in unprocessed objects}
library(here)

file_paths <- list.files(path = here::here("rds_unprocessed"), pattern = ".rds", full.names = TRUE)
file_names <- gsub(pattern = "_unprocessed.rds", replacement = "", x = basename(file_paths))

for(i in 1:length(file_names)){
  assign( file_names[i],readRDS(file_paths[i]))
}
```

```{r merge timepoint replicates}
ret = merge(E11, y = c(E12, E14.1, E14.2, E16, E18.2, E18.3, P0, P2.2, P5, P8.1, P14),
      add.cell.ids = c("E11", "E12", "E14.1", "E14.2", "E16", "E18.2", "E18.3", "P0", "P2.2", "P5", "P8.1", "P14"))
```

```{r add object metadata}
# add column for sample
ret$sample <- plyr::mapvalues(
  x = ret$orig.ident, 
  from = c("E11", "E12", "E14.1", "E14.2", "E16", "E18.2", "E18.3", "P0", "P2", "P5", "P8.1", "P14"), 
  to = c("E11", "E12", "E14.1", "E14.2", "E16", "E18.2", "E18.3", "P0", "P2.2", "P5", "P8.1", "P14")
)

# add column for timepoint
ret$timepoint <- plyr::mapvalues(
  x = ret$orig.ident, 
  from = c("E11", "E12", "E14.1", "E14.2", "E16", "E18.2", "E18.3", "P0", "P2", "P5", "P8.1", "P14"), 
  to = c("E11", "E12", "E14", "E14", "E16", "E18", "E18", "P0", "P2", "P5", "P8", "P14")
)

ret <- JoinLayers(ret)

saveRDS(ret, file = "complete_dev_ref_unprocessed.rds")
```

```{r Run normalization}
ret <- readRDS(file = "complete_dev_ref_unprocessed.rds")

library(glmGamPoi)

# SCT normalization
ret <- SCTransform(ret, vars.to.regress="percent.mt", conserve.memory = TRUE)

# save object
saveRDS(ret, "complete_dev_ref_normalized.rds")
```


```{r Process object}
# load object
ret <- readRDS("complete_dev_ref_normalized.rds")

# process
ret <- RunPCA(ret)
ret <- RunUMAP(ret, dims = 1:30, return.model=TRUE) ## adding return.model to use as reference for FindTransferAnchorts
ret <- FindNeighbors(ret, reduction = "pca", dims = 1:30)
ret <- FindClusters(ret, resolution = 0.4)

saveRDS(ret, file = "complete_dev_ref_processed.rds")
```

```{r Generate PCA and UMAP plots}
# PCA plots
DimPlot(ret, reduction = "pca")
ggsave(file= "complete_dev_ref_pca.pdf")

DimPlot(ret, reduction = "pca", split.by = "timepoint", ncol = 4)
ggsave(file= "complete_dev_ref_pca_bytimepoint.pdf")

ElbowPlot(ret, ndims=30)
ggsave(file= "complete_dev_ref_elbow.pdf")

# UMAPs
DimPlot(ret, group.by = "timepoint")
ggsave(file= "complete_dev_ref_umap_groupbytimepoint.pdf")

DimPlot(ret, split.by = "timepoint")
ggsave(file= "complete_dev_ref_umap_splitbytimepoint.pdf", width = 25)

DimPlot(ret, group.by = "sample", split.by = "timepoint")
ggsave(file= "complete_dev_ref_umap_groupbysample-splitbytimepoint.pdf", width = 25)

DimPlot(ret, group.by = "seurat_clusters")
ggsave(file= "complete_dev_ref_umap_byclusters.pdf")

DimPlot(ret, group.by = "seurat_clusters", split.by = "timepoint")
ggsave(file= "complete_dev_ref_umap_groupbyclusters_splitbytimepoint.pdf")
```

# Cluster Identification
```{r plot cell class marker expression}
# load library
library(stringr)

# list of genes taken from published EThomas 2019 code
# modified genes that don"t have the same name in mouse as in human
rods = c("PROM1", "CRX", "RCVRN", "OTX2", "RHO", "NR2E3", "NRL")
mg = c("SLC1A3", "SLN", "RLBP1", "C130071C03Rik", "SOX2", "NFIA", "CRYM", "CLU")
cones = c("ARR3", "PROM1", "CRX", "GNAT2", "OPN1SW", "PDE6H")
hc = c("ONECUT1", "ONECUT2", "ONECUT3", "TFAP2B", "LHX1")
bc = c("VSX1", "VSX2", "OTX2", "GRM6", "PRKCA", "LHX4", "PROX1", "PCP2", "PCP4", "TRPM1")
ac = c("SLC6A9", "GAD1", "SLC32A1", "TFAP2B", "GAD2", "SLC18A3", "LHX9", "TFAP2A", "TFAP2C")
rgc = c("POU4F2", "RBPMS", "NEFM", "GAP43", "POU4F1", "ELAVL4")
rpe = c("BEST1", "RPE65", "TIMP3", "TRPM1")
prog = c("VIM", "SOX2", "SFRP2", "MKI67", "UBE2C", "FGF15", "CCND1", "ID3", "PAX6", "VSX2", "LHX2", "HES1", "HES5")
prog2 = c("CCND1", "OLIG2", "NEUROG2")
prbcpre = c("VXN", "RXRG", "ATOH7", "PRDM1", "GADD45G", "DCT", "LHX4", "OTX2")
early = c("FGF15", "SFRP2", "BASP1", "KPNA2", "KIF2A", "GADD45A", "SOX11", "ELAVL3", "GAL" )
late = c("CRYM", "CAR2", "NFIA", "NFIB", "NFIX", "OAZ1", "PEPB1", "CD9", "SAT1", "GADD45G", "RGS16", "PRDM1", "ASCL1")
t1 = c("ATOH7", "HES6", "DLL3")
t2 = c("PRDM13", "HSPB1")
t3 = c("DLL3", "OTX2", "FABP7")
prog_pre = c("GADD45A", "NEUROG2", "OLIG2")
neuro_ep = c("MITF", "CCND2", "MSX1", "H19")
astro = c("NGFR", "GFAP", "SOX2", "SLC1A3")
microglia = c("C1QA", "TYROBP", "CD74", "P2RY12", "TMEM119", "CD45", "CD11B")
endo = c("GJA4", "BCAM", "ESAM")

celltypes = c("rods", "mg", "cones", "hc", "bc", "ac", "rgc", "rpe", "prog", "prbcpre", "prog2", "early", "late", "t1", "t2", "t3", "prog_pre", "neuro_ep", "astro", "microglia", "endo")

for (type in celltypes){
  obj <- get(type)
  genes = str_to_title(obj)

  DotPlot(ret, features = genes, group.by = "seurat_clusters") + RotatedAxis()
  ggsave(file = paste0("ret_dotplot_", type, ".pdf"))
}
```

```{r find cluster markers}
# Re-normalize data
DefaultAssay(ret) = "RNA"
ret = NormalizeData(ret)

markers <- FindAllMarkers(ret, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
write.csv(markers, file= "complete_dev_ref_clustermarkers.csv")
```

```{r assign celltypes based on cell class markers}
ret$celltype <-  plyr::mapvalues(ret$seurat_clusters,
                                 from = c(as.character(0:25)),
                                 to = c("Early_Progenitors", "Late_Progenitors", "Developing_Rods", "Late_Progenitors", "Early_Progenitors", "HC_AC_RGC_Precursors", "Developing_Rods", "Developing_Cones", "Retinal_Ganglion_Cells", "HC_AC_Precursors", "Amacrine_Cells", "Rods", "Neurogenic_Precursors", "Early_Progenitors", "Neurogenic_Precursors", "Developing_Muller_Glia", "Developing_Bipolar_Cells", "Developing_AC_HC", "Muller_Glia", "Bipolar_Cells", "Neuroepithelial_Progenitors", "Cones", "Neuroepithelial_Progenitors", "Neuroepithelial_Progenitors", "Amacrine_Cells", "Microglia"))

# create umaps of annotated celltypes
DimPlot(ret, group.by = "celltype")
ggsave(file= "complete_dev_ref_annotated_umap_bycelltype.pdf")

DimPlot(ret, group.by = "celltype", split.by = "timepoint")
ggsave(file= "complete_dev_ref_annotated_umap_bycelltype-splitbytimepoint.pdf", width = 20)

# save object
saveRDS(ret, file = "complete_dev_ref_annotated.rds")
```



# Dot Plot of miR9 host gene expression
```{r mir9 expression across time}
# load in object
ret <- readRDS(file = "complete_dev_ref_annotated.rds")

# normalize data
DefaultAssay(ret) = "RNA"
ret <- NormalizeData(ret)

# order by timepoint
ret$timepoint <- factor(ret$timepoint, levels = c("E11", "E12", "E14", "E16", "E18", "P0", "P2", "P5", "P8", "P14"))

# plot mir9-hg expression
DotPlot(ret, features = c("Mir9-3hg", "C130071C03Rik", "Gm3764"), group.by = "timepoint", cols = "Blues") + coord_flip() + RotatedAxis() + scale_color_distiller(direction = 0)
ggsave(file = "ret_dotplot_mir9expression_acrosstime.pdf", height = 3.5, width = 8)

```