---
title: "mir9enhancerKO_mice"
author: "LuLu Callies"
date: "2025-04-20"
output: html_document
---

# Setup

```{r load libraries}
library(Seurat)
library(SeuratData)
library(sctransform)
library(patchwork)
library(dplyr)
library(ggplot2)
library(harmony)
```

```{r Adjust the maximum size of global objects}
# (this may need to be increased later)
options(future.globals.maxSize = 200 * 1024^3)
```


# Load in data

```{r Read the 10x files}
#wildtype
eWT1.data <- Read10X_h5('WT1_cb_output_filtered_seurat.h5') # Cellbender model never completely converged for ELBO
eWT2.data <- Read10X_h5('WT2_cb_output_filtered_seurat.h5') # Cellbender model never completely converged for ELBO
eWT3.data <- Read10X_h5('WT3_cb_output_filtered_seurat.h5')
eKO1.data <- Read10X_h5('KO1_cb_output_filtered_seurat.h5') # Cellbender model never completely converged for ELBO
eKO2.data <- Read10X_h5('KO2_cb_output_filtered_seurat.h5')
eKO3.data <- Read10X_h5('KO3_cb_output_filtered_seurat.h5')
```

# Create Seurat objects
```{r Create Seurat objects}
#wildtype
eWT1 <- CreateSeuratObject(counts = eWT1.data, project = "eWT1", min.cells = 3, min.features = 200)
eWT2 <- CreateSeuratObject(counts = eWT2.data, project = "eWT2", min.cells = 3, min.features = 200)
eWT3 <- CreateSeuratObject(counts = eWT3.data, project = "eWT3", min.cells = 3, min.features = 200)
#knockout
eKO1 <- CreateSeuratObject(counts =  eKO1.data, project = "eKO1", min.cells = 3, min.features = 200)
eKO2 <- CreateSeuratObject(counts = eKO2.data, project = "eKO2", min.cells = 3, min.features = 200)
eKO3 <- CreateSeuratObject(counts =  eKO3.data, project = "eKO3", min.cells = 3, min.features = 200)
```

```{r add mitochondrial percentages}
#wildtype
eWT1[["percent.mt"]] <- PercentageFeatureSet(eWT1, pattern = "^mt-")
eWT2[["percent.mt"]] <- PercentageFeatureSet(eWT2, pattern = "^mt-")
eWT3[["percent.mt"]] <- PercentageFeatureSet(eWT3, pattern = "^mt-")
#knockout
eKO1[["percent.mt"]] <- PercentageFeatureSet(eKO1, pattern = "^mt-")
eKO2[["percent.mt"]] <- PercentageFeatureSet(eKO2, pattern = "^mt-")
eKO3[["percent.mt"]] <- PercentageFeatureSet(eKO3, pattern = "^mt-")
```

```{r Visualize QC metrics}
# (will split by 'orig.ident')
samples <- c('eWT1', 'eWT2', 'eWT3', 'eKO1', 'eKO2', 'eKO3')

for (x in samples){
  obj <- get(x)
  
  VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
  ggsave(file= paste0(x, '_qc.pdf'))
}
```

```{r Subset the data based on QC metric cutoffs}
#wildtype
eWT1 <- subset(eWT1, subset = nCount_RNA > 300 & nCount_RNA < 10000 & percent.mt < 5)
eWT2 <- subset(eWT2, subset = nCount_RNA > 300 & nCount_RNA < 10000 & percent.mt < 5)
eWT3 <- subset(eWT3, subset = nCount_RNA > 300 & nCount_RNA < 10000 & percent.mt < 5)
#knockout
eKO1 <- subset(eKO1, subset = nCount_RNA > 300 & nCount_RNA < 10000 & percent.mt < 5)
eKO2 <- subset(eKO2, subset = nCount_RNA > 300 & nCount_RNA < 10000 & percent.mt < 5)
eKO3 <- subset(eKO3, subset = nCount_RNA > 300 & nCount_RNA < 10000 & percent.mt < 5)
```

```{r merge into single object}
enh <- merge(eWT1, y = c(eWT2, eWT3, eKO1, eKO2, eKO3),
             add.cell.ids = c('eWT1', 'eWT2', 'eWT3', 'eKO1', 'eKO2', 'eKO3'))
```

```{r Visualize QC metrics}
# (will split by 'orig.ident')
VlnPlot(enh, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'enh_QC.pdf')
```

```{r add metadata}
enh$genotype <- plyr::mapvalues(
  x = enh$orig.ident, 
  from = c('eWT1', 'eWT2', 'eWT3', 'eKO1', 'eKO2', 'eKO3'), 
  to = c('WT','WT', 'WT', 'KO', 'KO','KO')
)

```


```{r Save merged Seurat objects}
saveRDS(enh, file = "enh.rds")
```

# Initial Analysis

```{r load in data}
enh <- readRDS("enh.rds")
```

```{r Run SCTransform and PCA}
# Normalize data by batch
enh <- JoinLayers(enh)
enh[['RNA']] <- split(enh[['RNA']], f = enh$orig.ident)

#Run normalization
enh <- SCTransform(enh, vars.to.regress = "percent.mt", verbose = FALSE)
enh <- RunPCA(enh, verbose = FALSE)
```

```{r Visualize samples in PCA space}
DimPlot(enh, reduction = 'pca')
ggsave(file= 'enh_pca.pdf')

DimPlot(enh, reduction = 'pca', split.by = 'orig.ident', ncol = 5)
ggsave(file= 'enh_pca_bysample.pdf')

ElbowPlot(enh, ndims=30)
ggsave(file= 'enh_elbow.pdf')
```

```{r Generate/visualize UMAP}
enh <- RunUMAP(enh, dims = 1:30, verbose = FALSE)

DimPlot(enh)
ggsave(file= 'enh_umap.unintegrated_bysample.pdf')

DimPlot(enh, group.by = 'genotype')
ggsave(file= 'enh_umap.unintegrated_groupbygenotype.pdf')

DimPlot(enh, group.by = 'genotype', split.by = 'genotype')
ggsave(file= 'enh_umap.unintegrated_splitbygenotype.pdf')

DimPlot(enh, group.by = 'orig.ident')
ggsave(file= 'enh_umap.unintegrated_groupbysample.pdf')

DimPlot(enh, split.by = 'orig.ident')
ggsave(file= 'enh_umap.unintegrated_splitbysample.pdf')
```

No need to use any sort of integration since the PCAs and UMAPs all appear to overlap and look similar

```{r finish inital processing}
# finish analysis
enh <- FindNeighbors(enh, dims = 1:30)
enh <- FindClusters(enh, resolution = 0.4)
enh <- RunUMAP(enh, dims = 1:30)
```

```{r visualize cluster umap}
DimPlot(enh)
ggsave(file= 'enh_clusters_bysample.pdf')

DimPlot(enh, split.by = 'orig.ident')
ggsave(file= 'enh_clusters_splitbysample.pdf')
```

```{r Save object}
DefaultAssay(enh) = 'RNA'
enh <- JoinLayers(enh)
saveRDS(enh, file = "enh_clusters.rds")
```

# Cluster ID

```{r reload object}
#reload
enh <- readRDS('enh_clusters.rds')
```

```{r Re-normalize}
#Set default assay to RNA, then re-normalize (SCT assay is good for data visualization but not quantitative comparisons, according to Satija lab)
DefaultAssay(enh) = 'RNA'
enh = NormalizeData(enh)
```

```{r look at QC by cluster}
VlnPlot(enh, features = c('nFeature_RNA', 'nCount_RNA', 'percent.mt'), pt.size=0)+NoLegend()
ggsave(file= 'enh_cluster_qc.pdf', width=20, height=10)
```

## Find clustermarkers

```{r find cluster markers}
markers <- FindAllMarkers(enh, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
write.csv(markers, file= 'enh_clustermarkers.csv')
```



## Automated cluster ID -- scMCA

```{r load libraries}
library(SingleCellExperiment)
library(scran)
library(scater)
library(dplyr)
library(scMCA)
library(Seurat)
library(ggplot2)
library(stringr)
```

```{r Adjust size of objects}
# Adjust the maximum size of global objects (this may need to be increased later)
options(future.globals.maxSize = 200 * 1024^3)
```

```{r load in data}
#data should be a SingleCellExperiment object, of normalized log counts
#load in Seurat object, with samples GEX-all and subset
enh <- readRDS('enh_clusters.rds')

#retrieve normalized logcounts
DefaultAssay(enh) = 'RNA'
enh = NormalizeData(enh)

# Retrieve or set data in an expression matrix ('counts', 'data', and 'scale.data')
mat <- GetAssayData(object = enh, layer = "counts")
```

```{r Run scMCA}
# scMCA has two parameters , single cell expression matrix(scdata) and the number of most similar cell types
set.seed(1234)
mca_result <- scMCA(scdata =  mat, numbers_plot = 3)
save(mca_result,file = "enh_mca_result.RData")
#load results
load('enh_mca_result.RData')
```

cors_matrix: Pearson correlation coefficient matrix of each cell and cell type.
top_cors: equals to numbers_plot
scMCA: the most relevant cell type for each query cell
scMCA_probility: the top n relevant cell types for each query cell

```{r Pull out relevant cell types}
# Extract the top probable
p <- mca_result$scMCA_probility
names(p) <- c("Cell","Cell_type", "Score") 

# For each cell select only those that have to do with retinas or at least 300 cells had that cell type as top one
freq_cell <- data.frame(table(p$Cell_type))
freq_cell <- freq_cell[stringr::str_detect( freq_cell$Var1,'Retina') | freq_cell$Freq>300, ] 


# Save the relevant cell types
relevant_cell_types <- freq_cell$Var1

# Check the cells wich top 10 cell clusters are not "relative to this study"
p <- p %>% group_by(Cell) %>% mutate(has_cluster= any(Cell_type%in%relevant_cell_types) )
ignore_cells <- unique(p$Cell[d$has_cluster==F])
length(ignore_cells)/length(unique(p$Cell))
```


```{r Extract cluster IDs}
ignore_p <- p[p$Cell%in%ignore_cells,]
ignore_p$Cell_type <- NA
ignore_p$Score <- 0
ignore_p$has_cluster <- F
ignore_p <- unique(ignore_p)

# Select the final selection of cells and their group
mca_labels <- p %>% group_by(Cell) %>% dplyr::filter(Cell_type%in%relevant_cell_types) %>% dplyr::filter(Score==max(Score)) %>% as.data.frame()
mca_labels <- rbind(mca_labels,ignore_p)
mca_labels$Cell <- as.character(mca_labels$Cell)
rownames(mca_labels) <- mca_labels$Cell

# extract just the new cluster ids from mca pipeline
mca_clusters <- as.data.frame(mca_labels$Cell_type)
rownames(mca_clusters) <- rownames(mca_labels)
colnames(mca_clusters) <- 'celltype'
```

```{r replace mca names}
# string replace mca cluster names with shortened versions
mca_clusters$mca_clusters <- case_when(
  mca_clusters$mca_clusters == "Rods(Retina)" ~ "Rods",
  mca_clusters$mca_clusters == "Cones(Retina)" ~ "Cones",
  mca_clusters$mca_clusters == "Amacrine cell_Tfap2b_high(Retina)" ~ "Amacrine_cells",
  mca_clusters$mca_clusters == "Rgs16 neurons(Arc-Me)" ~ "Rgs16",
  mca_clusters$mca_clusters == "Pericytes(Retina)" ~ "Pericytes",
  mca_clusters$mca_clusters == "Horizontal cells(Retina)" ~ "Horizontal_cells",
  mca_clusters$mca_clusters == "Retinal ganglion cells(Retina)" ~ "Retinal_ganglion_cells",
  mca_clusters$mca_clusters == "Amacrine cells_Rims1_high(Retina)" ~ "Amacrine_cells",
  mca_clusters$mca_clusters == "Microglia(Retina)" ~ "Microglia",
  mca_clusters$mca_clusters == "Muller glia(Retina)" ~ "Muller_glia",
  mca_clusters$mca_clusters == "Fibroblasts(Retina)" ~ "Fibroblasts",
  mca_clusters$mca_clusters == "Astrocytes(Retina)" ~ "Astrocytes",
  mca_clusters$mca_clusters == "Vascular endothelium(Retina)" ~ "Vascular_endothelium",
  mca_clusters$mca_clusters == "Amacrine cell_Cplx2_high(Retina)" ~ "Amacrine_cells",
  mca_clusters$mca_clusters == "Amacrine cell_Lamp5_high(Retina)" ~ "Amacrine_cells",
  mca_clusters$mca_clusters == "Bipolar cell_Gsg1_high(Retina)" ~ "Bipolar_cells",
  mca_clusters$mca_clusters == "Bipolar cell_Vsx1_high(Retina)" ~ "Bipolar_cells",
  mca_clusters$mca_clusters == "Bipolar cell_Lhx4_high(Retina)" ~ "Bipolar_cells",
  mca_clusters$mca_clusters == "Bipolar cell_Slit2_high(Retina)" ~ "Bipolar_cells",
  mca_clusters$mca_clusters == "Bipolar cells_Pcp2_high(Retina)" ~ "Bipolar_cells"
)
```

```{r add metadata to Seurat object}
enh <- readRDS('enh_clusters.rds')
enh <- AddMetaData(enh, mca_clusters)
head(enh@meta.data)
```

```{r save object}
saveRDS(enh, 'enh_mca.rds')
# this annotated object submitted to GEO as 'enh.rds'
```


```{r umap by mca clusters}
DimPlot(enh, group.by = 'celltype')
ggsave(file= 'enh_umap_celltype.pdf', height=7, width=15, dpi=72)

DimPlot(enh, label=TRUE, group.by = 'celltype') + NoLegend()
ggsave(file= 'enh_umap_celltype_labeled.pdf', height=7, width=15, dpi=72)
```

```{r make umaps with custom palette}
# make capitalization match palette
enh$celltype <- plyr::mapvalues(
  x = enh$celltype, 
  from = c(unique(enh$celltype)), 
  to = c('Cones', 'Amacrine_Cells', 'Bipolar_Cells', 'Muller_Glia', 'Rods', 'Retinal_Ganglion_Cells', 'Horizontal_Cells', unique(enh$mca_clusters)[8:14])
)

palette3 <- c("Amacrine_Cells" = "#332288", "Developing_AC_HC" = "#117733", "Developing_Cones" = "#44AA99", "Developing_Rods" = "#CC6677" , "Early_Progenitors" = "#666633", "HC_AC_Precursors" = "#DDCC77", "HC_AC_RGC_Precursors" = "#33BBEE", "Late_Progenitors" = "#882255", "Neurogenic_Precursors" = "#AA4499", "Retinal_Ganglion_Cells" = "#EE7733", "Bipolar_Cells" = "#66CCEE", "Cones" = "#228833", "Horizontal_Cells" = "#CCBB44", "Muller_Glia" = "#EE6677", "Rods" = "#AA3377")

enh <- subset(enh, subset = !(is.na(celltype)))

DimPlot(enh, group.by = 'celltype', cols = palette3)
ggsave(file= 'enh_umap_celltype_palette3.pdf', height=7, width=15, dpi=72)

DimPlot(enh, label=TRUE, group.by = 'celltype', cols = palette3) + NoLegend()
ggsave(file= 'enh_umap_celltype_labeled_palette3.pdf', height=7, width=15, dpi=72)
```


```{r differential expression using default presto}
library(presto)
library(Seurat)
library(edgeR)

# load object
enh <- readRDS('enh_mca.rds')
# need to re-normalize data because its best to use RNA assay for differential expression
#Set default assay to RNA, then re-normalize (SCT assay is good for data visualization but not quantitative comparisons, according to Satija lab)
DefaultAssay(enh) = 'RNA'
enh = NormalizeData(enh)

# create list of celltypes
celltypes <- c(unique(enh$celltype))
# need to filter out one celltype because it is only present in WT population
celltypes <- celltypes[1:7]

for (i in celltypes){
  cell <- as.character(i)
  
  Idents(enh) = 'celltype'
  subclass <- subset(enh, idents = cell)
  
  DefaultAssay(subclass) = 'RNA'
  
  # remove lowly expressed genes in celltype
  # use edgeR function which filters genes with min count in min number of samples
  counts <- Seurat2PB(subclass, sample = 'genotype', cluster = "celltype")
  keep <- edgeR::filterByExpr(counts) # use lenient parameters
  keep <- names(keep[keep == TRUE])
  subclass <- DietSeurat(subclass, layers = 'data', features = keep, assays = 'RNA')

  
  # set idents so DGE performed based on genotype
  Idents(subclass) = 'genotype'

  de = wilcoxauc(subclass, group_by = 'genotype', assay = 'data', seurat_assay = 'RNA')
  # filter data table to just KO
  de.ko <- de %>%
           filter(group == 'KO') %>%
           arrange(desc(auc))

  # save results
  write.csv(de.ko, file = paste0('de_presto_enhko_', cell, '_edgeRfilterbyExp.csv'))
}
```

```{r pseudobulk dge using presto}
# pseudobulk analysis which includes all celltypes lumped together

counts <- Seurat2PB(enh, sample = 'genotype', cluster = 'celltype')
keep <- edgeR::filterByExpr(counts) # use lenient parameters
keep <- names(keep[keep == TRUE])
keep <- DietSeurat(enh, layers = 'data', features = keep, assays = 'RNA')
  
# run analysis
de = wilcoxauc(enh, group_by = 'genotype', assay = 'data', seurat_assay = 'RNA')

# filter data to just knockout and order
de.ko <- de %>%
         filter(group == 'KO') %>%
         arrange(desc(auc))

# save results
write.csv(de.ko, file = 'de_enh_presto_pseudobulk_filterbyexp.csv')
```


```{r cell counts}
write.csv(table(enh$orig.ident, enh$celltype), file = 'enh_cellcounts_bycelltype.csv')
```

```{r generate plots of mir9 hg expression}
enh <- readRDS('enh_mca.rds')

DefaultAssay(enh) = 'RNA'
enh = NormalizeData(enh)

# generate violin plot of muller glia WT and KO
mg <- subset(enh, subset = celltype == 'Muller_glia')
mg$genotype <- factor(mg$genotype, levels = c('WT', 'KO'))

VlnPlot(mg, features = 'C130071C03Rik', pt.size = 0, group.by = 'genotype', cols = c('gray40', 'lightgray')) +
  # Add mean as a large, red diamond point
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4,
    color = "black"
  ) 

ggsave(file = 'violinplot_mg_mir9-2hg_expression.pdf', height = 10)


wt <- subset(enh, subset = genotype == 'WT')
DotPlot(wt, features = 'C130071C03Rik', cols = 'Blues', group.by = 'celltype', split.by = 'celltype') +
  coord_flip() +
  scale_color_distiller(direction = 0) +
  RotatedAxis()
```

```{r generate list of mir9 predicted targets}
library(readxl)

# import predicted mir9 targets for miR9-3p and miR9-5p, combine targets to make list of all mir9 targets
targets1 <- data.frame(read_xlsx('/data/hps/assoc/private/cherrylab/user/kcalli/miRNA_targets/TargetScan8.0__miR-9-5p.Mouse.predicted_targets.xlsx'))
targets2 <- data.frame(read_xlsx('/data/hps/assoc/private/cherrylab/user/kcalli/miRNA_targets/TargetScan8.0__miR-9-3p.Mouse.predicted_targets.xlsx'))
targets <- c(targets1$Target.gene, targets2$Target.gene)
targets <- unique(targets)
```


```{r create volcano plots for all cell classes}
library(stringr)
library(EnhancedVolcano)
library(dplyr)

targets <- str_to_title(targets)

celltypes <- unique(enh$celltype)
# remove celltypes with few cells/DEGs
celltypes <- celltypes[1:7]

for (i in celltypes){
  
  clust <- as.character(i)
  
  #output of Seurat's differential expression, subset by KO
  input = read.csv(paste0('de_presto_enhko_', clust, '_edgeRfilterbyExp.csv'))
  input$feature <- str_to_title(input$feature)
  
  labels <- filter(input, input$padj < 0.05) %>%
    filter(abs(logFC) > 0.1375) %>%
    arrange(padj)

  #Manipulate dataframes for downstream analysis
  input$mir9_target = ifelse(input$feature%in% targets, 'yes', 'no')
  input$mir9_up = ifelse(input$mir9_target == 'yes' & abs(input$logFC) >=0.137504, 'yes', 'no')
  targs3 = c('C130071C03Rik') 
  input$linc = ifelse(input$feature %in% targs3, 'yes', 'no')
  # modified mir9_up so that it shows both upregulated and downregulated mir9 targets
  # can change it back to upregulated only if you get rid of absolute value around log2FC
  
  #Set color and label parameters for volcano plot
  keyvals = ifelse(input$mir9_up == 'yes', '#7D3C98',  ifelse(input$linc == 'yes', '#F4D03F', ifelse(input$padj > 0.05, 'gray29', ifelse(input$logFC <= 0.137504 & input$logFC >= -0.137504, '#21618C', '#B71C1C')))) 
  names(keyvals)[keyvals == '#7D3C98'] <- 'Predicted miR-9-2 Target'
  names(keyvals)[keyvals == '#B71C1C'] <- 'Sig. p-value & logFC'
  names(keyvals)[keyvals == '#21618C'] <- 'Sig. p-value only'
  names(keyvals)[keyvals == '#F4D03F'] <- 'miR9-2 host gene'
  names(keyvals)[keyvals == 'gray29'] <- 'NS'
  
  #Define lists of genes to be labeled in plot
  input$mir9_label = ifelse(input$mir9_target == 'yes' & abs(input$logFC) >=0.26303, 'yes', 'no')
  mir9 = input[input$mir9_label == 'yes',]
  targs2 = as.character(mir9$feature)[1:20]
  targs3 = c('C130071C03Rik')
  targs4 = labels$feature[1:20]
  targs = c(targs2, targs3, targs4)

#Generate labeled volcano plot, with genes of interest labeled, colored, and increased in size
  volcano <- EnhancedVolcano(input,
                             lab = as.character(input$feature),
                              x = 'logFC',
                              y = 'padj',
                              xlab = bquote(~Log[2]~ 'fold change'),
                              xlim = c((min(input$logFC)-0.1), (max(input$logFC)+0.1)),
                              title = "",
                              subtitle = "",
                              selectLab = targs,
                              FCcutoff = 0.137504,
                              pCutoff = 0.05,
                              colAlpha = 3/4,
                              colCustom = keyvals,
                              labSize = 4,
                              pointSize = c(ifelse(input$endo == 'yes', 2, ifelse(input$mir9_up == 'yes', 2, 1))),
                              drawConnectors = T,
                              legendPosition = 'right',
                              legendLabSize = 12,
                              legendIconSize = 4.0,
                              max.overlaps = Inf,
                             arrowheads = FALSE
                             )                


ggsave(volcano, filename = paste0('enh_volcanoplot_', clust, '_edgeRfilterbyExp_2026-02.pdf'), width = 10)
}
```



```{r}
#output of Seurat's differential expression, subset by KO
  input = read.csv('de_enh_presto_pseudobulk_filterbyexp.csv')
  input$feature <- str_to_title(input$feature)
  
  labels <- filter(input, input$padj < 0.05) %>%
    filter(abs(logFC) > 0.1375) %>%
    arrange(padj)

  #Manipulate dataframes for downstream analysis
  input$mir9_target = ifelse(input$feature%in% targets, 'yes', 'no')
  input$mir9_up = ifelse(input$mir9_target == 'yes' & abs(input$logFC) >=0.137504, 'yes', 'no')
  targs3 = c('C130071C03Rik') 
  input$linc = ifelse(input$feature %in% targs3, 'yes', 'no')
  # modified mir9_up so that it shows both upregulated and downregulated mir9 targets
  # can change it back to upregulated only if you get rid of absolute value around log2FC
  
  #Set color and label parameters for volcano plot
  keyvals = ifelse(input$mir9_up == 'yes', '#7D3C98',  ifelse(input$linc == 'yes', '#F4D03F', ifelse(input$padj > 0.05, 'gray29', ifelse(input$logFC <= 0.137504 & input$logFC >= -0.137504, '#21618C', '#B71C1C')))) 
  names(keyvals)[keyvals == '#7D3C98'] <- 'Predicted miR-9-2 Target'
  names(keyvals)[keyvals == '#B71C1C'] <- 'Sig. p-value & logFC'
  names(keyvals)[keyvals == '#21618C'] <- 'Sig. p-value only'
  names(keyvals)[keyvals == '#F4D03F'] <- 'miR9-2 host gene'
  names(keyvals)[keyvals == 'gray29'] <- 'NS'
  
  #Define lists of genes to be labeled in plot
  input$mir9_label = ifelse(input$mir9_target == 'yes' & abs(input$logFC) >=0.26303, 'yes', 'no')
  mir9 = input[input$mir9_label == 'yes',]
  targs2 = as.character(mir9$feature)[1:20]
  targs3 = c('C130071C03Rik')
  targs4 = labels$feature[1:20]
  targs = c(targs2, targs3, targs4)

#Generate labeled volcano plot, with genes of interest labeled, colored, and increased in size
  volcano <- EnhancedVolcano(input,
                             lab = as.character(input$feature),
                              x = 'logFC',
                              y = 'padj',
                              xlab = bquote(~Log[2]~ 'fold change'),
                              xlim = c((min(input$logFC)-0.1), (max(input$logFC)+0.1)),
                              title = "",
                              subtitle = "",
                              selectLab = targs,
                              FCcutoff = 0.137504,
                              pCutoff = 0.05,
                              colAlpha = 3/4,
                              colCustom = keyvals,
                              labSize = 4,
                              pointSize = c(ifelse(input$endo == 'yes', 2, ifelse(input$mir9_up == 'yes', 2, 1))),
                              drawConnectors = T,
                              legendPosition = 'right',
                              legendLabSize = 12,
                              legendIconSize = 4.0,
                              max.overlaps = Inf,
                             arrowheads = FALSE
                             )                


ggsave(volcano, filename = 'enh_volcanoplot_pseudobulk_edgeRfilterbyExp_2026-02.pdf', width = 10)
```


# Generate cell class proportions chart with uniform palette

```{r load libraries}
library(Seurat)
library(plyr)
library(dplyr)
```


```{r define uniform color palette used for all plots}
palette3 <- c("Amacrine_Cells" = "#332288", "Developing_AC_HC" = "#117733", "Developing_Cones" = "#44AA99", "Developing_Rods" = "#CC6677" , "Early_Progenitors" = "#666633", "HC_AC_Precursors" = "#DDCC77", "HC_AC_RGC_Precursors" = "#33BBEE", "Late_Progenitors" = "#882255", "Neuroepithelial_Progenitors" = "lightgrey", "Neurogenic_Precursors" = "#AA4499", "Retinal_Ganglion_Cells" = "#EE7733", "Bipolar_Cells" = "#66CCEE", "Cones" = "#228833", "Horizontal_Cells" = "#CCBB44", "Muller_Glia" = "#EE6677", "Rods" = "#AA3377")
```


```{r generate stacked bar plots}
# load in data
enh <- readRDS('enh_mca.rds')

enh$celltype <- plyr::mapvalues(
  x = enh$celltype, 
  from = c(unique(enh$celltype)), 
  to = c('Cones', 'Amacrine_Cells', 'Bipolar_Cells', 'Muller_Glia', 'Rods', 'Retinal_Ganglion_Cells', 'Horizontal_Cells', unique(enh$celltype)[8:14])
)

enh <- subset(enh, subset = !(enh$celltype %in% c('Rgs16', 'Fibroblasts', 'Astrocytes', 'Pericytes', 'Vascular_endothelium', 'Microglia')))

# create proportion table
enh.proportions <- prop.table(table(enh$celltype, enh$orig.ident), 2)

# reshape data into dataframe
enh.comparison <- data.frame(enh.proportions)
colnames(enh.comparison) <- c('Celltype', 'Sample', 'Proportion')
enh.comparison$genotype <- plyr::mapvalues(enh.comparison$Sample,
                                 from = c('eWT1', 'eWT2', 'eWT3', 'eKO1', 'eKO2', 'eKO3'),
                                 to = c('WT', 'WT', 'WT', 'KO', 'KO', 'KO'))

# plot data
library(ggplot2)

enh.comparison$genotype <- factor(enh.comparison$genotype, levels=rev(c('WT', 'KO')))
enh.comparison$Sample <- factor(enh.comparison$Sample, levels=rev(c('eWT1', 'eWT2', 'eWT3', 'eKO1', 'eKO2', 'eKO3')))

# by sample
enh.comparison %>%
  ggplot(aes(x=Sample, y=Proportion, fill=Celltype))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  RotatedAxis() +
  coord_flip() +
  scale_fill_manual(values = palette3)

ggsave(filename = 'enh_stackedbar_cellclassproportions_bysample-celltype_horizontal.pdf')

#by genotype
enh.comparison %>%
  ggplot(aes(x=genotype, y=Proportion, fill=Celltype))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  RotatedAxis() +
  coord_flip() +
  scale_fill_manual(values = palette3)

ggsave(filename = 'enh_stackedbar_cellclassproportions_bygeno-celltype_horizontal.pdf')
```

```{r}
write.csv(table(enh$celltype, enh$orig.ident), file = 'enh_cellcounts_celltypebysample.csv')

write.csv(table(enh$orig.ident), file = 'enh_cellspersample.csv')
```


```{r Generate umap by genotype with uniform palette}
# load in data
enh <- readRDS('enh_mca.rds')

enh$celltype <- plyr::mapvalues(
  x = enh$celltype, 
  from = c(unique(enh$celltype)), 
  to = c('Cones', 'Amacrine_Cells', 'Bipolar_Cells', 'Muller_Glia', 'Rods', 'Retinal_Ganglion_Cells', 'Horizontal_Cells', unique(enh$celltype)[8:14])
)

enh <- subset(enh, subset = !(enh$celltype %in% c('Rgs16', 'Fibroblasts', 'Astrocytes', 'Pericytes', 'Vascular_endothelium', 'Microglia')))

DimPlot(enh, reduction = 'umap', group.by = 'celltype', cols = palette3)
ggsave(file = 'umap_bycelltype_palette3.pdf')

enh$genotype <- factor(enh$genotype, levels = c('WT', 'KO'))

DimPlot(enh, reduction = 'umap', group.by = 'celltype', split.by = 'genotype', cols = palette3)
ggsave(filename = 'umap_bycelltype_splitbygeno_palette3.pdf', width = 14, height = 7)
```










