---
title: "p0retina_seurat5"
author: "LuLu Callies"
output: html_document
date: "2024-09-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r load libraries}
library(Seurat)
library(SeuratData)
library(sctransform)
library(patchwork)
library(dplyr)
library(ggplot2)
library(harmony)
library(readxl)
library(presto)
```

```{r Adjust the maximum size of global objects}
# (this may need to be increased later)
options(future.globals.maxSize = 200 * 1024^3)
```

# Load in data

```{r Read the 10x files}
#wildtype
WTP0F1.data <- Read10X_h5(filename = 'WTP0F1_output_filtered_seurat.h5')
WTP0F2.data <- Read10X_h5(filename = 'WTP0F2_output_filtered_seurat.h5')
WTP0M1.data <- Read10X_h5(filename = 'WTP0M1_output_filtered_seurat.h5')
WTP0M2.data <- Read10X_h5(filename = 'WTP0M2_output_filtered_seurat.h5')
#knockout
KOP0F1.data <- Read10X_h5(filename = 'KOP0F1_output_filtered_seurat.h5')
KOP0F2.data <- Read10X_h5(filename = 'KOP0F2_output_filtered_seurat.h5')
KOP0M1.data <- Read10X_h5(filename = 'KOP0M1_output_filtered_seurat.h5')
KOP0M2.data <- Read10X_h5(filename = 'KOP0M2_output_filtered_seurat.h5')
```

# Create Seurat objects
```{r Create Seurat objects}
#wildtype
WTP0F1 <- CreateSeuratObject(counts = WTP0F1.data, project = "WTP0F1", min.cells = 3, min.features = 200)
WTP0F2 <- CreateSeuratObject(counts = WTP0F2.data, project = "WTP0F2", min.cells = 3, min.features = 200)
WTP0M1 <- CreateSeuratObject(counts = WTP0M1.data, project = "WTP0M1", min.cells = 3, min.features = 200)
WTP0M2 <- CreateSeuratObject(counts = WTP0M2.data, project = "WTP0M2", min.cells = 3, min.features = 200)
#knockout
KOP0F1 <- CreateSeuratObject(counts =  KOP0F1.data, project = "KOP0F1", min.cells = 3, min.features = 200)
KOP0F2 <- CreateSeuratObject(counts = KOP0F2.data, project = "KOP0F2", min.cells = 3, min.features = 200)
KOP0M1 <- CreateSeuratObject(counts =  KOP0M1.data, project = "KOP0M1", min.cells = 3, min.features = 200)
KOP0M2 <- CreateSeuratObject(counts = KOP0M2.data, project = "KOP0M2", min.cells = 3, min.features = 200)
```

```{r add mitochondrial percentages}
#wildtype
WTP0F1[["percent.mt"]] <- PercentageFeatureSet(WTP0F1, pattern = "^mt-")
WTP0F2[["percent.mt"]] <- PercentageFeatureSet(WTP0F2, pattern = "^mt-")
WTP0M1[["percent.mt"]] <- PercentageFeatureSet(WTP0M1, pattern = "^mt-")
WTP0M2[["percent.mt"]] <- PercentageFeatureSet(WTP0M2, pattern = "^mt-")
#knockout
KOP0F1[["percent.mt"]] <- PercentageFeatureSet(KOP0F1, pattern = "^mt-")
KOP0F2[["percent.mt"]] <- PercentageFeatureSet(KOP0F2, pattern = "^mt-")
KOP0M1[["percent.mt"]] <- PercentageFeatureSet(KOP0M1, pattern = "^mt-")
KOP0M2[["percent.mt"]] <- PercentageFeatureSet(KOP0M2, pattern = "^mt-")
```

```{r Visualize QC metrics}
# (will split by 'orig.ident')

samples <- c('WTP0M1', 'WTP0M2', 'WTP0F1', 'WTP0F2', 'KOP0M1', 'KOP0M2', 'KOP0F1', 'KOP0F2')

for (sample in samples){
  obj <- get(sample)
  
  VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
  ggsave(file = paste0(sample, "_qc.pdf"))
}
```

```{r Subset the data based on QC metric cutoffs}
#wildtype
WTP0F1 <- subset(WTP0F1, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
WTP0F2 <- subset(WTP0F2, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
WTP0M1 <- subset(WTP0M1, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
WTP0M2 <- subset(WTP0M2, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
#knockout
KOP0F1 <- subset(KOP0F1, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
KOP0F2 <- subset(KOP0F2, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
KOP0M1 <- subset(KOP0M1, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
KOP0M2 <- subset(KOP0M2, subset = nCount_RNA > 1000 & nCount_RNA < 15000 & percent.mt < 5)
```

```{r Merge into single Seurat object}
p0=merge(WTP0F1, y = c(WTP0F2, WTP0M1, WTP0M2, KOP0F1, KOP0F2, KOP0M1, KOP0M2), 
          add.cell.ids = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'))
```

```{r Add sample and condition information to metadata}
# add column for batch
p0$batch <- plyr::mapvalues(
  x = p0$orig.ident,
  from = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'),
  to = c('batch1', 'batch1', 'batch2', 'batch2', 'batch1', 'batch1', 'batch2', 'batch2')
)

# add column for sample
p0$sample <- plyr::mapvalues(
  x = p0$orig.ident, 
  from = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'), 
  to = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2')
)

# order sample ids by genotype and sex
Idents(p0) = 'sample'
Idents(p0) <- factor(Idents(p0), levels = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'))

# add column for timepoint
p0$timepoint <- plyr::mapvalues(
  x = p0$orig.ident, 
  from = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'), 
  to = c('P0','P0','P0','P0', 'P0', 'P0', 'P0', 'P0')
)

# add column for genotype
p0$genotype <- plyr::mapvalues(
  x = p0$orig.ident, 
  from = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'), 
  to = c('WT', 'WT', 'WT','WT', 'KO', 'KO', 'KO', 'KO')
)

# order genotype
Idents(p0) = 'genotype'
Idents(p0) <- factor(Idents(p0), levels = c('WT', 'KO'))

# add column for sex
p0$sex <- plyr::mapvalues(
  x = p0$orig.ident,
  from = c('WTP0F1', 'WTP0F2', 'WTP0M1', 'WTP0M2', 'KOP0F1', 'KOP0F2', 'KOP0M1', 'KOP0M2'),
  to = c('female', 'female', 'male', 'male', 'female', 'female', 'male', 'male')
)
```

```{r Visualize QC metrics}
# order samples by genotype and sex
p0$sample <- factor(p0$sample, levels = c('WTP0M1', 'WTP0M2', 'WTP0F1', 'WTP0F2', 'KOP0M1', 'KOP0M2', 'KOP0F1', 'KOP0F2'))

VlnPlot(p0, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE, group.by = 'sample')
ggsave(file= 'p0_qc.pdf')
```



```{r Save merged Seurat objects}
saveRDS(p0, file = "p0.rds")
```

# Initial Analysis

```{r load in data}
p0 <- readRDS("p0.rds")
```


```{r Run SCTransform and PCA}
# split sample by batch for normalization
p0 <- JoinLayers(p0)
p0[['RNA']] <- split(p0[['RNA']], f = p0$batch)

# run normalization and PCA
p0 <- SCTransform(p0, vars.to.regress = "percent.mt", verbose = FALSE)
p0 <- RunPCA(p0)
```

## Without Harmony

```{r Visualize samples in PCA space}
DimPlot(p0, reduction = 'pca')
ggsave(file= 'p0_pca.pdf')

DimPlot(p0, reduction = 'pca', split.by = 'sample', ncol = 4)
ggsave(file= 'p0_pca_bysample.pdf')

DimPlot(p0, reduction = 'pca', split.by = 'batch', ncol = 4)
ggsave(file= 'p0_pca_bybatch.pdf')

DimPlot(p0, reduction = 'pca', split.by = 'sex', ncol = 4)
ggsave(file= 'p0_pca_bysex.pdf')

ElbowPlot(p0, ndims=30)
ggsave(file= 'p0_elbow.pdf')
```

```{r Generate/visualize UMAP}
p0 <- RunUMAP(p0, dims = 1:30)

DimPlot(p0, group.by = 'sample')
ggsave(file= 'p0_umap_bysample.pdf')

DimPlot(p0, group.by = 'genotype')
ggsave(file= 'p0_umap_bygenotype.pdf')

DimPlot(p0, group.by = 'genotype', split.by = 'genotype')
ggsave(file= 'p0_umap_splitbygenotype.pdf')

DimPlot(p0, group.by = 'genotype', split.by = 'batch')
ggsave(file= 'p0_umap_bygenotype-batch.pdf')

DimPlot(p0, group.by = 'sex', split.by = 'sex')
ggsave(file= 'p0_umap_splitbysex.pdf')
```


## With Harmony

```{r Run harmony for comparison}
# PCA Analysis with harmony integration
p0_harmony <- RunHarmony(p0, group.by.vars = 'sample', assay.use = 'SCT') %>% 
  RunUMAP(dims = 1:50, reduction = 'harmony', reduction.name = 'umap.harmony')

saveRDS(p0_harmony, file = "p0_harmony.rds")
```

```{r Visualize UMAP}
DimPlot(p0_harmony, reduction = 'umap.harmony', group.by = 'genotype')
ggsave(file= 'p0_harmony_umap_bygenotype.pdf')

DimPlot(p0_harmony, reduction = 'umap.harmony', group.by = 'genotype', split.by = 'genotype')
ggsave(file= 'p0_harmony_umap_splitbygenotype.pdf')

DimPlot(p0_harmony, reduction = 'umap.harmony', split.by = 'batch')
ggsave(file= 'p0_harmony_umap_splitbybatch.pdf')
```

## Finish workflow

Proceeding with harmony. Samples cluster separately without batch correction.

```{r Finish standard Seurat workflow}
#(FindNeighbors, FindClusters)
p0_harmony <- FindNeighbors(p0_harmony, reduction = 'harmony', dims = 1:30, verbose=FALSE)
p0_harmony <- FindClusters(p0_harmony, resolution = 0.4) 

DimPlot(p0_harmony, reduction='umap.harmony', label=TRUE)+NoLegend()
ggsave(file= 'p0_umap_harmonybycluster.pdf')

DimPlot(p0_harmony, reduction='umap.harmony', label=FALSE, split.by = 'genotype')
ggsave(file= 'p0_umap_harmonybycluster_splitbygenotype.pdf')

DimPlot(p0_harmony, reduction='umap.harmony', label=FALSE, split.by = 'sex')
ggsave(file= 'p0_umap_harmonybycluster_splitbysex.pdf')

DimPlot(p0_harmony, reduction='umap.harmony', label=FALSE, split.by = 'batch')
ggsave(file= 'p0_umap_harmonybycluster_splitbybatch.pdf')
```

For Seurat v5, make sure to rejoin layers once integration is complete. Rejoin before any differential expression analysis

```{r Save object}
# Join layers post-integration
DefaultAssay(p0_harmony) = 'RNA'
p0_harmony <- JoinLayers(p0_harmony)

# save object
saveRDS(p0_harmony, file = "p0_clusters.rds")
```

# Cluster ID

```{r reload object}
#reload
p0 <- readRDS('p0_clusters.rds')
```

```{r Re-normalize}
#Set default assay to RNA, then re-normalize (SCT assay is good for data visualization but not quantitative comparisons, according to Satija lab)
DefaultAssay(p0) = 'RNA'

# Normalize based on batch and then rejoin layers
p0[['RNA']] <- split(p0[['RNA']], f = p0$batch)
p0 = NormalizeData(p0)
p0 <- JoinLayers(p0)
```

```{r look at QC by cluster}
VlnPlot(p0, features = c('nFeature_RNA', 'nCount_RNA', 'percent.mt'), pt.size=0)+NoLegend()
ggsave(file= 'p0_cluster_qc.pdf', width=20, height=10)
```

## Find clustermarkers and check cluster expression

```{r find cluster markers}
markers <- FindAllMarkers(p0, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
write.csv(markers, file= 'p0_clustermarkers.csv')
```


```{r Look at cell class marker expression across clusters}
DefaultAssay(p0) = 'RNA'

genes <- c('C130071C03Rik', 'Rho', 'Sag', 'Gnat1', 'Nr2e3', 'Nrl', 'Gnat2', 'Rorb', 'Lhx1', 'Onecut3', 'Vsx2', 'Slc1a3', 'Tfa2b', 'Gad1', 'Gad2', 'Meis2', 'Rbpms', 'Sox9', 'Rlbp1', 'Otx2', 'Crx', 'Mki67', 'Atoh7')

for (gene in genes){
  FeaturePlot(p0, reduction = 'umap.harmony', features = gene, split.by = 'genotype')
  ggsave(paste0('p0_harmony_global_', gene, '_featureplot.pdf', width=15))
}
```



# Map reference labels onto dataset
```{r load in data}
p0 <- readRDS(file = 'p0_clusters.rds')
DefaultAssay(p0) = 'SCT'

# load in references
ret_ref <- readRDS('complete_dev_ref_annotated.rds')
DefaultAssay(ret_ref) = 'SCT'
```


```{r subset reference and map to seurat object}
# create subset of reference corresponding to timepoints harvested
ret_sub <- subset(ret_ref, subset = timepoint %in% c('E14', 'E16', 'E18', 'P0', 'P2'))
ret_sub <- RunUMAP(ret_sub, reduction = 'pca', dims = 1:30, return.model = TRUE)

DimPlot(ret_sub, reduction = 'umap', group.by = 'timepoint')
ggsave(file = 'dev.E14.P2-ref_umap_bytimepoint.tif')

DimPlot(ret_sub, reduction = 'umap', group.by = 'celltype')
ggsave(file = 'dev.E14.P2-ref_umap_bycelltype.tif')

# compute anchors between reference and dataset
anchors <- FindTransferAnchors(
  reference = ret_sub,
  query = p0,
  normalization.method = "SCT",
  reference.reduction = "pca",
  dims = 1:50
)

# map labels onto dataset
p0 <- MapQuery(
  anchorset = anchors,
  query = p0,
  reference = ret_sub,
  refdata = 'celltype',
  reference.reduction = "pca", 
  reduction.model = "umap"
)

# save mapped object
saveRDS(p0, file = 'p0_E14-P2_refmapping.rds')
# this annotated object sbmitted to GEO as 'p0.rds'
```

```{r plot results of mapping}
# create harmony umap of predicted celltypes
DimPlot(p0, reduction = 'umap.harmony', group.by = 'predicted.id')
ggsave('p0_umap.harmony_predicted.id.tif')

DimPlot(p0, reduction = 'umap.harmony', group.by = 'predicted.id', split.by = 'genotype')
ggsave('p0_umap.harmony_predicted.id_bygenotype.pdf', width = 14)
```

# Customize UMAPs

```{r create UMAPs with uniform color palette}
# load in object
p0 <- readRDS(file = 'p0_E14-P2_refmapping.rds')

# set custom colors
palette <- c("Amacrine_Cells" = "#332288", "Developing_AC_HC" = "#117733", "Developing_Cones" = "#44AA99", "Developing_Rods" = "#CC6677" , "Early_Progenitors" = "#666633", "HC_AC_Precursors" = "#DDCC77", "HC_AC_RGC_Precursors" = "#33BBEE", "Late_Progenitors" = "#882255", "Neurogenic_Precursors" = "#AA4499", "Retinal_Ganglion_Cells" = "#EE7733", "Bipolar_Cells" = "#66CCEE", "Cones" = "#228833", "Horizontal_Cells" = "#CCBB44", "Muller_Glia" = "#EE6677", "Rods" = "#AA3377")

# create umaps for each genotype
genotypes <- c('WT', 'KO')

## subset out celltypes with small numbers
celltypes <- c("Amacrine_Cells", "Bipolar_Cells", "Cones", "Developing_Cones", "Horizontal_Cells", "Muller_Glia", "Retinal_Ganglion_Cells", "Rods", "Developing_Rods", "Early_Progenitors", "Late_Progenitors", "Neurogenic_Precursors", "Developing_AC_HC", "HC_AC_RGC_Precursors", "HC_AC_Precursors")
p0_sub <- subset(p0, subset = predicted.id %in% celltypes)


for (geno in genotypes){
  sub <- subset(p0_sub, subset = genotype == geno)

  DimPlot(sub, reduction='umap.harmony', group.by = 'predicted.id', raster=FALSE, cols = palette, label = TRUE, pt.size = 1, stroke = 0, shuffle = TRUE)
  ggsave(file= paste0('p0_', geno, '_umap_harmony_bypredictedid.pdf'), width = 10)
}
```


# Compare cell class proportions for KO vs WT
IDs based on predicted celltypes

```{r calculate and plot proportions}
# create dataframe of cell class proportions
p0.proportions <- prop.table(table(p0$predicted.id, p0$sample),2)

#reorganize data
library(reshape)
library(plyr)
p0.comparison <- melt(p0.proportions, id=c("X"))
colnames(p0.comparison) <- c('Celltype', 'Sample', 'Proportion')
p0.comparison$genotype <- plyr::mapvalues(p0.comparison$Sample,
                                 from = c('WTP0M1', 'WTP0M2', 'WTP0F1', 'WTP0F2', 'KOP0M1', 'KOP0M2', 'KOP0F1', 'KOP0F2'),
                                 to = c('WT', 'WT', 'WT', 'WT', 'KO', 'KO', 'KO', 'KO'))

p0.comparison$genotype <- factor(p0.comparison$genotype, levels=rev(c('WT', 'KO')))
p0.comparison$Sample <- factor(p0.comparison$Sample, levels=rev(c('WTP0M1', 'WTP0M2', 'KOP0M1', 'KOP0M2','WTP0F1', 'WTP0F2', 'KOP0F1', 'KOP0F2')))
```


```{r plot proportions}
# plot data
library(ggplot2)

# set color palette
palette <- c("Amacrine_Cells" = "#332288", "Developing_AC_HC" = "#117733", "Developing_Cones" = "#44AA99", "Developing_Rods" = "#CC6677" , "Early_Progenitors" = "#666633", "HC_AC_Precursors" = "#DDCC77", "HC_AC_RGC_Precursors" = "#33BBEE", "Late_Progenitors" = "#882255", "Neuroepithelial_Progenitors" = "lightgrey", "Neurogenic_Precursors" = "#AA4499", "Retinal_Ganglion_Cells" = "#EE7733", "Bipolar_Cells" = "#66CCEE", "Cones" = "#228833", "Horizontal_Cells" = "#CCBB44", "Muller_Glia" = "#EE6677", "Rods" = "#AA3377")


# stacked bar plots by sample
p0.comparison %>%
  ggplot(aes(x=Sample, y=Proportion, fill=Celltype))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  RotatedAxis() +
  coord_flip() +
  scale_fill_manual(values = palette)

ggsave(filename = 'p0_stackedbar_cellclassproportions_predictedcelltypebysample.pdf')

# stacked bar plots by genotype
p0.comparison %>%
  ggplot(aes(x=genotype, y=Proportion, fill=Celltype))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  RotatedAxis() +
  coord_flip() +
  scale_fill_manual(values = palette)

ggsave(filename = 'p0_stackedbar_cellclassproportions_predictedcelltypebygenotype.pdf')
```

## Run differential abundance analysis for cell class proportions
Based on EdgeR differential abundance analysis

```{r create tables of cell counts}
p0 <- readRDS(file = 'p0_E14-P2-refmapping.rds')

# create table of celltypes by sample
cellcounts <- table(p0$predicted.id, p0$sample)

# Filter out celltypes with low counts
mincells <- 100L
cellcounts <- cellcounts[rowSums(cellcounts) > mincells,]
```

```{r use edgeR to calculate statistics}
# load in packages
library(edgeR)

dP0 <- DGEList(cellcounts,
                samples = data.frame(row.names = colnames(cellcounts),
                                     Genotype = factor(substring(colnames(cellcounts), 1, 2), levels = c("WT", "KO")))
designP0 <- model.matrix(~ Genotype, dP0$samples)

dP0 <- estimateDisp(dP0, designP0, trend = "none")
summary(dP0$common.dispersion)
fit.P0 <- glmQLFit(dP0, designP0, robust = TRUE, abundance.trend = FALSE)
summary(fit.P0$var.prior)
summary(fit.P0$df.prior)

tiff("AllSamples/P0_QL_dispersion_2024-07-11.tiff", compression = "lzw", res = 200, width = 7 * 200, height = 5 * 200)
plotQLDisp(fit.P0, cex = 1)
dev.off()

# statistical significance for celltypes
# significant change in developing rods and amacrines
res.P0 <- glmQLFTest(fit.P0, coef = 2)
p0.stat <- topTags(res.P0)
write.csv(p0.stat, file = 'p0_proportiontable_statistics.csv')
```

# Cell Cycle Scoring

```{r load in data and required lists}
# load libraries
library(Seurat)
library(dplyr)
library(ggplot2)
library(harmony)

# load in lists of cell cycle genes
library(gprofiler2)
mmus_s = gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m = gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

# load in object
p0 <- readRDS(file = 'p0_E14-P2-refmapping.rds')
```


```{r assign cell cycle scores}
p0 <- CellCycleScoring(p0, s.features = mmus_s, g2m.features = mmus_g2m, set.ident = TRUE)
```

```{r look at results of scoring}
# look at scores plotted on umap
DimPlot(p0, reduction = 'umap.harmony', group.by = 'Phase')
ggsave(file = 'umap_p0_cellcyclephase.tif')

DimPlot(p0, reduction = 'umap.harmony', group.by = 'Phase', split.by = 'genotype')
ggsave(file = 'umap_p0_cellcyclephase_bygenotype.tif', width = 14)
```

```{r look at phase proportions}
# calculate proportion table of cell cycle phase by sample
proportions <- prop.table(table(p0$Phase, p0$sample),2) # 2 indicates to create by columns

# reshape data and annotate data
comparison <- data.frame(proportions)
colnames(comparison) <- c('Phase', 'Sample', 'Proportion')
comparison$Genotype <- plyr::mapvalues(comparison$Sample,
                                 from = c('WTP0M1', 'WTP0M2', 'WTP0F1', 'WTP0F2', 'KOP0M1', 'KOP0M2', 'KOP0F1', 'KOP0F2'),
                                 to = c('WT', 'WT', 'WT', 'WT', 'KO', 'KO', 'KO', 'KO'))
# plot proportions
library(ggplot2)

comparison$Genotype <- factor(comparison$Genotype, levels=c('WT', 'KO'))
comparison$Sample <- factor(comparison$Sample, levels=c('WTP0M1', 'WTP0M2', 'KOP0M1', 'KOP0M2','WTP0F1', 'WTP0F2', 'KOP0F1', 'KOP0F2'))

# by sample
comparison %>%
  ggplot(aes(x=Sample, y=Proportion, fill=Phase))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  RotatedAxis()

ggsave(filename = 'p0_stackedbar_cellcycleproportions_bysample.tif')

#by genotype
comparison %>%
  ggplot(aes(x=Genotype, y=Proportion, fill=Phase))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  RotatedAxis()

ggsave(filename = 'p0_stackedbar_cellcycleproportions_bygeno.tif', width = 14)
```

```{r calculate statistical significance of phase differences}
# create table of celltypes by sample
cellcounts <- table(p0$Phase, p0$sample)

# use edgeR to calculate statistics
# load in packages
library(edgeR)

dP0 <- DGEList(cellcounts,
                samples = data.frame(row.names = colnames(cellcounts),
                                     Genotype = factor(substring(colnames(cellcounts), 1, 2), levels = c("WT", "KO")),
                                     Sex = substring(colnames(cellcounts), 5, 5)))
designP0 <- model.matrix(~ Genotype, dP0$samples)

dP0 <- estimateDisp(dP0, designP0, trend = "none")
summary(dP0$common.dispersion) # all 0.0692
fit.P0 <- glmQLFit(dP0, designP0, robust = TRUE, abundance.trend = FALSE)
summary(fit.P0$var.prior) # all 21.11
summary(fit.P0$df.prior) # all 1.43

tiff("p0_QL_dispersion_cellcyclephase.tiff", compression = "lzw", res = 200, width = 7 * 200, height = 5 * 200)
plotQLDisp(fit.P0, cex = 1)
dev.off()

# statistical significance for cell cycle phase
res.P0 <- glmQLFTest(fit.P0, coef = 2)
P0.stat <- topTags(res.P0)
write.csv(P0.stat, file = 'p0_cellcyclephase_proportiontable_statistics.csv')
```
