---
title: "wk3retina_Seurat5"
output: html_document
date: "2024-09-18"
author: "LuLu Callies"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(Seurat)
library(dplyr)
library(ggplot2)
library(harmony)
```

```{r Adjust the maximum size of global objects}
# (this may need to be increased later)
options(future.globals.maxSize = 200 * 1024^3)
```

# Load in data

```{r Read the 10x files}
#wildtype
WT3WM1.data <- Read10X_h5('WT3WM1_output_filtered_seurat.h5')
WT3WM3.data <- Read10X_h5('WT3WM3_output_filtered_seurat.h5')
WT3WF1.data <- Read10X_h5('WT3WF1_output_filtered_seurat.h5')
WT3WF2.data <- Read10X_h5('WT3WF2_output_filtered_seurat.h5')
WT3WF3.data <- Read10X_h5('WT3WF3_output_filtered_seurat.h5')
#knockout
KO3WM1.data <- Read10X_h5('KO3WM1_output_filtered_seurat.h5')
KO3WM3.data <- Read10X_h5('KO3WM3_output_filtered_seurat.h5')
KO3WF1.data <- Read10X_h5('KO3WF1_output_filtered_seurat.h5')
KO3WF2.data <- Read10X_h5('KO3WF2_output_filtered_seurat.h5')
KO3WF3.data <- Read10X_h5('KO3WF3_output_filtered_seurat.h5')
```

# Create Seurat objects
```{r Create Seurat objects}
# Create Seurat objects
#wildtype
WT3WM1 <- CreateSeuratObject(counts = WT3WM1.data, project = "WT3WM1", min.cells = 3, min.features = 200)
WT3WM3 <- CreateSeuratObject(counts = WT3WM3.data, project = "WT3WM3", min.cells = 3, min.features = 200)
WT3WF1 <- CreateSeuratObject(counts = WT3WF1.data, project = "WT3WF1", min.cells = 3, min.features = 200)
WT3WF2 <- CreateSeuratObject(counts = WT3WF2.data, project = "WT3WF2", min.cells = 3, min.features = 200)
WT3WF3 <- CreateSeuratObject(counts = WT3WF3.data, project = "WT3WF3", min.cells = 3, min.features = 200)
#knockout
KO3WM1 <- CreateSeuratObject(counts = KO3WM1.data, project = "KO3WM1", min.cells = 3, min.features = 200)
KO3WM3 <- CreateSeuratObject(counts = KO3WM3.data, project = "KO3WM3", min.cells = 3, min.features = 200)
KO3WF1 <- CreateSeuratObject(counts = KO3WF1.data, project = "KO3WF1", min.cells = 3, min.features = 200)
KO3WF2 <- CreateSeuratObject(counts = KO3WF2.data, project = "KO3WF2", min.cells = 3, min.features = 200)
KO3WF3 <- CreateSeuratObject(counts = KO3WF3.data, project = "KO3WF3", min.cells = 3, min.features = 200)
```

```{r add mitochondrial percentages}
#wildtype
WT3WM1[["percent.mt"]] <- PercentageFeatureSet(WT3WM1, pattern = "^mt-")
WT3WM3[["percent.mt"]] <- PercentageFeatureSet(WT3WM3, pattern = "^mt-")
WT3WF1[["percent.mt"]] <- PercentageFeatureSet(WT3WF1, pattern = "^mt-")
WT3WF2[["percent.mt"]] <- PercentageFeatureSet(WT3WF2, pattern = "^mt-")
WT3WF3[["percent.mt"]] <- PercentageFeatureSet(WT3WF3, pattern = "^mt-")
#knockout
KO3WM1[["percent.mt"]] <- PercentageFeatureSet(KO3WM1, pattern = "^mt-")
KO3WM3[["percent.mt"]] <- PercentageFeatureSet(KO3WM3, pattern = "^mt-")
KO3WF1[["percent.mt"]] <- PercentageFeatureSet(KO3WF1, pattern = "^mt-")
KO3WF2[["percent.mt"]] <- PercentageFeatureSet(KO3WF2, pattern = "^mt-")
KO3WF3[["percent.mt"]] <- PercentageFeatureSet(KO3WF3, pattern = "^mt-")
```

```{r Visualize QC metrics}
# (will split by 'orig.ident')

#wildtype
VlnPlot(WT3WM1, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'WT3WM1_qc.pdf')

VlnPlot(WT3WM3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'WT3WM3_qc.pdf')

VlnPlot(WT3WF1, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'WT3WF1_qc.pdf')

VlnPlot(WT3WF2, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'WT3WF2_qc.pdf')

VlnPlot(WT3WF3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'WT3WF3_qc.pdf')

#knockout
VlnPlot(KO3WM1, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'KO3WM1_qc.pdf')

VlnPlot(KO3WM3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'KO3WM3_qc.pdf')

VlnPlot(KO3WF1, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'KO3WF1_qc.pdf')

VlnPlot(KO3WF2, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'KO3WF2_qc.pdf')

VlnPlot(KO3WF3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'KO3WF3_qc.pdf')
```

```{r Subset the data based on QC metric cutoffs}
#wildtype
WT3WM1 <- subset(WT3WM1, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
WT3WM3 <- subset(WT3WM3, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
WT3WF1 <- subset(WT3WF1, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
WT3WF2 <- subset(WT3WF2, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
WT3WF3 <- subset(WT3WF3, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
#knockout
KO3WM1 <- subset(KO3WM1, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
KO3WM3 <- subset(KO3WM3, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
KO3WF1 <- subset(KO3WF1, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
KO3WF2 <- subset(KO3WF2, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
KO3WF3 <- subset(KO3WF3, subset = nCount_RNA > 500 & nCount_RNA < 100000 & percent.mt < 5)
```

```{r Merge into single Seurat object}
wk3=merge(WT3WM1, y = c(WT3WM3, WT3WF1, WT3WF2, WT3WF3, KO3WM1, KO3WM3, KO3WF1, KO3WF2, KO3WF3), 
          add.cell.ids = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'))
```

```{r Visualize QC metrics}
# (will split by 'orig.ident')
VlnPlot(wk3, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0, log = TRUE)
ggsave(file= 'wk3_cellbender_QC.pdf')
```

```{r Add sample and condition information to metadata}
wk3$batch <- plyr::mapvalues(
  x = wk3$orig.ident,
  from = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'),
  to = c('batch1', 'batch2', 'batch1', 'batch3', 'batch3', 'batch1', 'batch2', 'batch1', 'batch3', 'batch3')
)

wk3$sample <- plyr::mapvalues(
  x = wk3$orig.ident, 
  from = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'), 
  to = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3')
)

wk3$timepoint <- plyr::mapvalues(
  x = wk3$orig.ident, 
  from = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'), 
  to = c('3wk','3wk','3wk','3wk','3wk','3wk', '3wk', '3wk', '3wk', '3wk')
)

wk3$genotype <- plyr::mapvalues(
  x = wk3$orig.ident, 
  from = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'), 
  to = c('WT','WT', 'WT', 'WT','WT','KO','KO','KO','KO','KO')
)

wk3$sex <- plyr::mapvalues(
  x = wk3$orig.ident, 
  from = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'), 
  to = c('male', 'male', 'female', 'female', 'female', 'male', 'male', 'female', 'female', 'female')
)
```

```{r Save merged Seurat objects}
saveRDS(wk3, file = "wk3.rds")
```

# Initial Analysis

```{r load in data}
wk3 <- readRDS('wk3.rds')
```

# Without integration

```{r Run SCTransform and PCA}
# Split object by batches and analyze without integration
wk3 <- JoinLayers(wk3)

# Split into layers by batch
wk3[['RNA']] <- split(wk3[['RNA']], f = wk3$batch) 

# initial processing
wk3 <- SCTransform(wk3, vars.to.regress = "percent.mt", verbose = FALSE)
wk3 <- RunPCA(wk3, verbose = FALSE)
wk3 <- RunUMAP(wk3, dims = 1:30)
```


```{r Visualize pca without batch correction}
DimPlot(wk3, reduction = 'pca', group.by = 'sample')
ggsave(file= 'wk3_pca_unintegrated_bysample.pdf')

DimPlot(wk3, reduction = 'pca', group.by = 'batch', split.by = 'sample', ncol = 5)
ggsave(file= 'wk3_pca_unintegrated_bybatch-sample.pdf')

DimPlot(wk3, reduction = 'pca', group.by = 'batch')
ggsave(file= 'wk3_pca_unintegrated_bybatch.pdf')

ElbowPlot(wk3, ndims=30)
ggsave(file= 'wk3_elbow.pdf')
```


```{r Visualize umap without batch correction}
DimPlot(wk3, reduction = 'umap', group.by = 'sample')
ggsave(file= 'wk3_umap_unintegrated_bysample.pdf')

DimPlot(wk3, reduction = 'umap', group.by = 'batch', split.by = 'sample', ncol = 5)
ggsave(file= 'wk3_umap_unintegrated_bybatch-sample.pdf')

DimPlot(wk3, reduction = 'umap', group.by = 'batch')
ggsave(file= 'wk3_umap_unintegrated_bybatch.pdf')
```


## Harmony integration

```{r integrate with harmony}
# integrate layers
wk3_harmony <- IntegrateLayers(object = wk3, method = HarmonyIntegration, normalization.method = 'SCT', new.reduction = "harmony")

# finish analysis
wk3_harmony <- FindNeighbors(wk3_harmony, reduction = 'harmony', dims = 1:30)
wk3_harmony <- FindClusters(wk3_harmony, resolution = 0.4)
wk3_harmony <- RunUMAP(wk3_harmony, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")

# save harmony object
saveRDS(wk3_harmony, file = 'wk3_harmony.rds')
```

```{r Visualize umaps with harmony}
DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'sample')
ggsave(file= 'wk3_umap.harmony_integrated_bysample.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'batch', split.by = 'sample', ncol = 5)
ggsave(file= 'wk3_umap.harmony_integrated_bybatch-sample.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'batch')
ggsave(file= 'wk3_umap.harmony_integrated_bybatch.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'seurat_clusters')
ggsave(file= 'wk3_umap.harmony_integrated_bycluster.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'seurat_clusters', split.by = 'genotype')
ggsave(file= 'wk3_umap.harmony_integrated_bycluster-genotype.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'genotype')
ggsave(file= 'wk3_umap.harmony_integrated_groupbygenotype.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'genotype', split.by = 'genotype')
ggsave(file= 'wk3_umap.harmony_integrated_splitbygenotype.pdf')

DimPlot(wk3_harmony, reduction = 'umap.harmony', group.by = 'genotype', split.by = 'batch')
ggsave(file= 'wk3_umap.harmony_integrated_bygenotype-batch.pdf')
```

Proceeding with Harmony based on batch effects

```{r Save object}
# Rejoin layers before downstream analysis
DefaultAssay(wk3_harmony) = 'RNA'
wk3_harmony <- JoinLayers(wk3_harmony)

# save object
saveRDS(wk3_harmony, file = "wk3_clusters.rds")
```

# Cluster Identification - Clustermarkers

```{r ignoreoad object}
#ignoreoad
wk3 <- readRDS("wk3_clusters.rds")
```

```{r Re-normalize}
#Set default assay to RNA, then re-normalize (SCT assay is good for data visualization but not quantitative comparisons, according to Satija lab)
DefaultAssay(wk3) = 'RNA'
wk3[['RNA']] <- split(wk3[['RNA']], f = wk3$batch)
wk3 = NormalizeData(wk3)
wk3 <- JoinLayers(wk3)
```

```{r look at QC by cluster}
VlnPlot(wk3, features = c('nFeature_RNA', 'nCount_RNA', 'percent.mt'), pt.size=0)+NoLegend()
ggsave(file= 'wk3_qc_bycluster.pdf', width=20, height=10)
```

```{r find cluster markers}
markers <- FindAllMarkers(wk3, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
write.csv(markers, file= 'wk3_clustermarkers.csv')
```


#Automated cluster ID -- scMCA

```{r load libraries}
library(SingleCellExperiment)
library(scran)
library(scater)
library(dplyr)
library(scMCA)
library(Seurat)
library(ggplot2)
library(stringr)
```

```{r Adjust size of objects}
# Adjust the maximum size of global objects (this may need to be increased later)
options(future.globals.maxSize = 200 * 1024^3)
```

```{r load in data}
#data should be a SingleCellExperiment object, of normalized log counts
#load in Seurat object, with samples GEX-all and subset
wk3 <- readRDS('wk3_clusters.rds')

#retrieve normalized logcounts
DefaultAssay(wk3) = 'RNA'
wk3[['RNA']] <- split(wk3[['RNA']], f = wk3$batch)
wk3 = NormalizeData(wk3)
wk3 <- JoinLayers(wk3)

# Retrieve or set data in an expression matrix ('counts', 'data', and 'scale.data')
mat <- GetAssayData(object = wk3, layer = "counts")
```

```{r Run scMCA}
# scMCA has two parameters , single cell expression matrix(scdata) and the number of most similar cell types
set.seed(1234)
mca_result <- scMCA(scdata =  mat, numbers_plot = 3)
save(mca_result,file = "wk3_mca_result.RData")
#load results
load("wk3_mca_result.RData")
```

cors_matrix: Pearson correlation coefficient matrix of each cell and cell type.
top_cors: equals to numbers_plot
scMCA: the most relevant cell type for each query cell
scMCA_probility: the top n relevant cell types for each query cell

```{r Pull out relevant cell types}
# Extract the top probable
p <- mca_result$scMCA_probility
names(p) <- c("Cell","Cell_type", "Score") 

# For ech cell select only those that have to do with retinas or at least 300 cells had that cell type as top one!
freq_cell <- data.frame(table(p$Cell_type))
freq_cell <- freq_cell[stringr::str_detect( freq_cell$Var1,'Retina') | freq_cell$Freq>300, ] 

# Save the relevant cell types
relevant_cell_types <- freq_cell$Var1

# Check the cells wich top 10 cell clusters are not "relative to this study"
p <- p %>% group_by(Cell) %>% mutate(has_cluster= any(Cell_type%in%relevant_cell_types) )
ignore_cells <- unique(p$Cell[p$has_cluster==F])
length(ignore_cells)/length(unique(p$Cell))
# In this case, only 0.05% of cells had a type that was not relevant
```


```{r Extract cluster IDs}
relevant_p <- p[p$Cell%in%rougue_cells,]
relevant_p$Cell_type <- NA
relevant_p$Score <- 0
relevant_p$has_cluster <- F
relevant_p <- unique(relevant_p)

# Select the final selection of cells and their group
mca_labels <- p %>% group_by(Cell) %>% dplyr::filter(Cell_type%in%relevant_cell_types) %>% dplyr::filter(Score==max(Score)) %>% as.data.frame()
mca_labels <- rbind(mca_labels, relevant_p)
mca_labels$Cell <- as.character(mca_labels$Cell)
rownames(mca_labels) <- mca_labels$Cell

# extract just the new cluster ids from mca pipeline
celltype <- as.data.frame(mca_labels$Cell_type)
rownames(celltype) <- rownames(mca_labels)
colnames(celltype) <- 'celltype'
```

```{r replace mca names}
# string replace mca cluster names with shortened versions
celltype$celltype <- case_when(
  celltype$celltype == "Rods(Retina)" ~ "Rods",
  celltype$celltype == "Cones(Retina)" ~ "Cones",
  celltype$celltype == "Amacrine cell_Tfap2b_high(Retina)" ~ "Amacrine_cells",
  celltype$celltype == "Rgs16 neurons(Arc-Me)" ~ "Rgs16",
  celltype$celltype == "Pericytes(Retina)" ~ "Pericytes",
  celltype$celltype == "Horizontal cells(Retina)" ~ "Horizontal_cells",
  celltype$celltype == "Retinal ganglion cells(Retina)" ~ "Retinal_ganglion_cells",
  celltype$celltype == "Amacrine cells_Rims1_high(Retina)" ~ "Amacrine_cells",
  celltype$celltype == "Microglia(Retina)" ~ "Microglia",
  celltype$celltype == "Muller glia(Retina)" ~ "Muller_glia",
  celltype$celltype == "Fibroblasts(Retina)" ~ "Fibroblasts",
  celltype$celltype == "Astrocytes(Retina)" ~ "Astrocytes",
  celltype$celltype == "Vascular endothelium(Retina)" ~ "Vascular_endothelium",
  celltype$celltype == "Amacrine cell_Cplx2_high(Retina)" ~ "Amacrine_cells",
  celltype$celltype == "Amacrine cell_Lamp5_high(Retina)" ~ "Amacrine_cells",
  celltype$celltype == "Bipolar cell_Gsg1_high(Retina)" ~ "Bipolar_cells",
  celltype$celltype == "Bipolar cell_Vsx1_high(Retina)" ~ "Bipolar_cells",
  celltype$celltype == "Bipolar cell_Lhx4_high(Retina)" ~ "Bipolar_cells",
  celltype$celltype == "Bipolar cell_Slit2_high(Retina)" ~ "Bipolar_cells",
  celltype$celltype == "Bipolar cells_Pcp2_high(Retina)" ~ "Bipolar_cells"
)
```

```{r add metadata to Seurat object}
wk3 <- readRDS('wk3_clusters.rds')
wk3 <- AddMetaData(wk3, celltype)
head(wk3@meta.data)
```

```{r save object}
saveRDS(wk3, 'wk3_refmapping.rds')
# this annotated object submitted to GEO as 'wk3.rds'
```

```{r umap by celltype}
DimPlot(wk3, group.by = 'celltype')
ggsave(file= 'wk3_umap_celltype.pdf', height=7, width=15, dpi=72)

DimPlot(wk3, reduction='umap.harmony', label=TRUE, group.by = 'celltype')
ggsave(file= 'wk3_umap_celltype_labeled.pdf', height=7, width=15, dpi=72)

DimPlot(wk3, reduction='umap.harmony', label=TRUE, group.by = 'celltype', split.by = 'batch')
ggsave(file= 'wk3_umap_celltype-batch.pdf', height=7, width=15, dpi=72)
```


# Create custom UMAP

```{r create UMAP using standardized colors}
# load in object
wk3 <- readRDS(file = 'wk3_refmapping.rds')

# remap labels to match capitalization to palette
wk3$celltype <- plyr::mapvalues(
  x = wk3$celltype,
  from = c('Amacrine_Cells', 'Amacrine_cells','Bipolar_cells', 'Muller_glia', 'Horizontal_cells', 'Retinal_ganglion_cells'),
  to = c('Developing_Amacrine_Cells', 'Amacrine_Cells', 'Bipolar_Cells', 'Muller_Glia', 'Horizontal_Cells', 'Retinal_Ganglion_Cells')
)

### Generate WT and KO UMAPs using manual custom color palette ###
palette3 <- c("Amacrine_Cells" = "#332288", "Developing_AC_HC" = "#117733", "Developing_Cones" = "#44AA99", "Developing_Rods" = "#CC6677" , "Early_Progenitors" = "#666633", "HC_AC_Precursors" = "#DDCC77", "HC_AC_RGC_Precursors" = "#33BBEE", "Late_Progenitors" = "#882255", "Neurogenic_Precursors" = "#AA4499", "Retinal_Ganglion_Cells" = "#EE7733", "Bipolar_Cells" = "#66CCEE", "Cones" = "#228833", "Horizontal_Cells" = "#CCBB44", "Muller_Glia" = "#EE6677", "Rods" = "#AA3377")

# create umaps for each genotype
genotypes <- c('WT', 'KO')
## subset to only celltypes of interest
celltypes <- c("Amacrine_Cells", "Bipolar_Cells", "Cones", "Developing_Cones", "Horizontal_Cells", "Muller_Glia", "Retinal_Ganglion_Cells", "Rods", "Developing_Rods", "Early_Progenitors", "Late_Progenitors", "Neurogenic_Precursors", "Developing_AC_HC", "HC_AC_RGC_Precursors", "HC_AC_Precursors")

# numbering system:
# 1. Amacrine Cells 
# 2. Developing AC HC
# 3. Developing Cones
# 4. Developing Rods
# 5. Early Progenitors
# 6. HC AC Precursors
# 7. HC AC RGC Precursors
# 8. Late Progenitors
# 9. Neurogenic Precursors
# 10. Retinal Ganglion Cells
# 11. Bipolar Cells
# 12. Cones
# 13. Horizontal Cells
# 14. Muller Glia
# 15. Rods

wk3_sub <- subset(wk3, subset = celltype %in% celltypes)

for (geno in genotypes){
  sub <- subset(wk3_sub, subset = genotype == geno)

  DimPlot(sub, reduction='umap.harmony', group.by = 'celltype', cols = palette3, raster=TRUE)
  ggsave(file= paste0('wk3_', geno, '_umap_harmony_celltypes.pdf'), width = 10, dpi=300)
}
```

# Check for expected gene expression

### Standard cell class markers
```{r cell class marker expression}
DefaultAssay(wk3) = 'RNA'

genes <- c('Rho', 'Sag', 'Gnat1', 'Gnat2', 'Lhx1', 'Vsx1', 'Vsx2', 'Tfap2b', 'Rbpms', 'Rlbp1', 'Slc1a3', 'Sox9')

for gene (in genes){
  FeaturePlot(object = wk3, reduction = 'umap.harmony', features= gene, cols = c("lightgrey", "blue"), pt.size = 0.5, split.by = 'genotype')
  ggsave(paste0('wk3_featureplot_', gene, '_splitbygenotype.pdf'), height = 15, width = 15)
}

### LINC expression

```{r mir9 expression}
FeaturePlot(wk3, reduction = 'umap.harmony', features = 'C130071C03Rik', split.by = 'genotype')
ggsave('wk3_linc_featureplot_splitbygenotype.pdf', width=15)
```

# Cell class proportions

```{r create dataframe of cell class proportions}
# load in object
wk3 <- readRDS(file = 'wk3_refmapping.rds')

# create proportion table
wk3.proportions <- prop.table(table(wk3$celltype, wk3$sample), 2)

# reshape data into dataframe
wk3.comparison <- data.frame(wk3.proportions)
colnames(wk3.comparison) <- c('Celltype', 'Sample', 'Proportion')

# add genotype info
wk3.comparison$genotype <- plyr::mapvalues(wk3.comparison$Sample,
                                 from = c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3'),
                                 to = c('WT', 'WT', 'WT', 'WT', 'WT', 'KO', 'KO', 'KO', 'KO', 'KO'))
```

```{r plot proportions by sample}
# set colors
palette3 <- c("Amacrine_Cells" = "#332288", "Developing_AC_HC" = "#117733", "Developing_Cones" = "#44AA99", "Developing_Rods" = "#CC6677" , "Early_Progenitors" = "#666633", "HC_AC_Precursors" = "#DDCC77", "HC_AC_RGC_Precursors" = "#33BBEE", "Late_Progenitors" = "#882255", "Neurogenic_Precursors" = "#AA4499", "Retinal_Ganglion_Cells" = "#EE7733", "Bipolar_Cells" = "#66CCEE", "Cones" = "#228833", "Horizontal_Cells" = "#CCBB44", "Muller_Glia" = "#EE6677", "Rods" = "#AA3377")

wk3.comparison$genotype <- factor(wk3.comparison$genotype, levels=rev(c('WT', 'KO')))
wk3.comparison$Sample <- factor(wk3.comparison$Sample, levels=rev(c('WT3WM1', 'WT3WM3', 'WT3WF1', 'WT3WF2', 'WT3WF3', 'KO3WM1', 'KO3WM3', 'KO3WF1', 'KO3WF2', 'KO3WF3')))

# create stacked barchart
wk3.comparison %>%
  ggplot(aes(x=Sample, y=Proportion, fill=Celltype))+
  geom_bar(position = 'stack', stat = 'identity') +
  theme_classic() +
  scale_fill_manual(values = palette3) +
  coord_flip() +
  RotatedAxis()

ggsave(filename = 'wk3_stackedbar_cellclassproportions_bysample.pdf')
```


# Differential abundance analysis
Based on edgeR differential abundance analysis

```{r create tables of cell counts}
wk3 <- readRDS(file = 'wk3_refmapping.rds')

# create table of celltypes by sample
cellcounts <- table(wk3$celltype, wk3$sample)

# Filter out celltypes with low counts
mincells <- 100L
cellcounts <- cellcounts[rowSums(cellcounts) > mincells,]
```

```{r use edgeR to calculate statistics}
# load in packages
library(edgeR)

dwk3 <- DGEList(cellcounts,
                samples = data.frame(row.names = colnames(cellcounts),
                                     Genotype = factor(substring(colnames(cellcounts), 1, 2), levels = c("WT", "KO")),
                                     Sex = substring(colnames(cellcounts), 5, 5)))
designwk3 <- model.matrix(~ Genotype, dwk3$sample)

dwk3 <- estimateDisp(dwk3, designwk3, trend = "none")
summary(dwk3$common.dispersion)
fit.wk3 <- glmQLFit(dwk3, designwk3, robust = TRUE, abundance.trend = FALSE)
summary(fit.wk3$var.prior)
summary(fit.wk3$df.prior)

tiff("wk3_QL_dispersion.tiff", compression = "lzw", res = 200, width = 7 * 200, height = 5 * 200)
plotQLDisp(fit.wk3, cex = 1)
dev.off()

# statistical significance for celltypes
res.wk3 <- glmQLFTest(fit.wk3, coef = 2)
wk3.stat <- topTags(res.wk3)
write.csv(wk3.stat, file = 'wk3_proportiontable_statistics.csv')
```


